### 2016-08-26 Chapters/PaperSummary.tex
Work on the summary of papers

\epigraph{$\frac{1}{2}$ research and $\frac{1}{2}\,\,T\acute{\epsilon}\chi\nu\eta$

          $\frac{1}{2}$ observation, $\frac{1}{2}\,\,T\acute{\epsilon}\chi\nu\eta$

          $\frac{1}{2}$ training, $\frac{1}{2}\,\,T\acute{\epsilon}\chi\nu\eta$}{
  --- \textsc{Ezra Pound}, \textit{Canto LXXXV}}

In the following, I provide a brief summary of the motivations, contents and
conclusions of the papers this thesis is based on.
Figures and tables here reported are minimally modified versions taken from the published work,
captions are instead copied \emph{verbatim}.
In addition, I give a description of my contributions to the work leading to
the publication of each of the papers.
\todo[inline]{Write something along these lines: "My coauthors have all read and approved."}

@@

\section*{Software}

@@

\begin{figure}[!htb]
\centering
\scalebox{0.8}{\input{gfx/relapcm/algorithm.tikz}}
\caption[Modular approach to programming a \acs{PCM} functionality into an existing \acs{SCF} code.]{
  Schematic view of the implemented SCF algorithm. Computations/data in
  \textcolor{Blue}{blue} are on the \pcmsolver side, in
  \textcolor{Green}{green} on the \DIRAC side.
  }
\label{fig:algorithm}
\end{figure}

>>>>

% Very ugly
\epigraph{\textonehalf\,\,research\,\,\,\,\,\,\,\,\,\,\,\,and \textonehalf\,\,\,\,\,\,\,\,Tεχνη

          \textonehalf\,\,observation,\,\,\,\,\,\,\,\,\,\,\,\,\textonehalf\,\,\,\,\,\,\,\,Tεχνη

          \textonehalf\,\,training,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\textonehalf\,\,\,\,\,\,\,\,Tεχνη}{
  --- \textsc{Ezra Pound}, \textit{Canto LXXXV}}

This final Chapter provides a brief overview of the motivations, results and
conclusions of the papers this thesis is based on.
All publications included in this thesis have involved some programming effort.
I will also describe the principles we have striven to follow in developing our
\acs{PCM} software library and its management philosophy.
I will also give some details on the interfaces we have developed to different
quantum chemistry codes.
Finally, for each paper, I will give a description of my contributions to the
work leading to its publication.
\todo[inline]{Write something along these lines: "My coauthors have all read and approved."}

@@

\refstepcounter{dummy}
\addcontentsline{toc}{section}{Software}
\section*{Software}

\todo[inline]{Have a closer look at this. I think it needs to be updated for the discussion
on open-source software that went down on JPC Lett. some time ago.
Also, acronyms should be properly set up.}

A shorter version of this discussion has been accepted for publication in the
proceedings of the \emph{Producing High Performance and Sustainable Software
for Molecular Simulation} workshop held at the 2015 Supercomputing Conference.

We have created an open-source, standalone library to provide \acs{PCM} functionality across
altogether different quantum chemical program packages.
The library serves as our development platform for the \acs{PCM}.
Moreover, the interfaces developed to different quantum chemistry codes allow the exploration
of new methodologies, significantly cutting down development times.

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{notpretty.png}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\scalebox{0.8}{\input{gfx/algorithm.tikz}}
\caption[Modular approach to programming a \acs{PCM} functionality into an existing \acs{SCF} code.]{
  Schematic view of the implemented SCF algorithm. Computations/data in
  \textcolor{Blue}{blue} are on the \pcmsolver side, in
  \textcolor{Green}{green} on the \DIRAC side.
  }
\label{fig:algorithm}
\end{figure}

The growing complexity of quantum chemical program packages requires that an
appropriate strategy be devised to implement new features.
Scalability is of paramount importance, but it has become clear that
maintainability and extensibility of the code play an equally important role in
managing this growing complexity.~\autocite{Hatton1994-us, Hatton1997-ml,
Hatton1997-bb, Ioannidis2005-aj, Merali2010-uy, Prinz2011-kz, Wilson2014-vh}

The use of a \emph{modular programming paradigm} has been recognized as
beneficial in many other scientific computing contexts. New features are
isolated into libraries that can be accessed by host programs through a
well-defined \ac{API}.
Computational tasks are thus implemented into separate, independent and
interchangeable modules. These are \emph{developed} and \emph{tested}
independently of the particular host program in which they will be
used.~\cite{Dijkstra1968-vy, Parnas1972-im}
A well-defined \ac{API} clearly delimits the boundaries of the functionality
offered, thus forcing a programming style and standardisation of the
functionality, eventually.~\autocite{Reddy2011-sd}

When coupled with \emph{open-source} licensing, modularity is a proper
step towards ensuring reproducibility of results from scientific
simulation software \cite{Gezelter2015-gz, Krylov2015-fs, Jacob2016-oq}.
Open-source development can fully leverage the benefits
of widespread, cloud-based, free, code development services, such as
hosted \acp{DVCS}, \ac{CI}, code
coverage, static and dynamic code analyses, nightly regression testing,
public issue tracking, code review and so forth:
adoption of a modern code development workflow is easily within reach.

It is, of course, true that the above mentioned services are not
exclusive prerogatives of open-source projects.
However, an open review process of scientific software can often help
to establish
reproducibility, extensibility and sustainability of the software
ecosystem.
Open-source software, modular development of new functionalities
and full-fledged exploitation of \acsp{DVCS} ensure a much larger scientific
impact. Third-parties can easily contribute to
the project: by improving the documentation, by reporting bugs or
by actively extending the codebase with new functionality.

The description of chemical phenomena in complex environments, such as
solutions or proteins, is as challenging as it is important
for an accurate prediction of a wide range of properties.~\autocite{Reichardt2010-le}
  Quantum chemistry is nowadays a valuable
tool: its simulations can provide significant insight into complex
chemical phenomena.~\autocite{Nobel1998, Nobel2013}.
The major challenge lies in providing a faithful, realistic, yet cost-effective
description of the environment of the molecule:
a molecule in solution can be represented by a quantum mechanical model
comprising a great number of atoms, thousands or more. Such a simulation would,
however, be impossible to carry out with the current state-of-the-art
computational resources.
Models must be devised to overcome the dimensionality ``disease'' and
the idea of introducing different approximation levels for different
parts of the system under study is particularly appealing, as it helps
our chemical understanding. This is the idea at the basis of
\emph{multiscale} models.

\begin{figure}[!htb]
  \centering
  \includegraphics[width=.45\textwidth]{multiscale.jpg}
\end{figure}

\emph{Continuum} (or \emph{implicit}) models, deal explicitly
only with the degrees of freedom associated with the solute, while
replacing the solvent with a structureless continuum characterized by
its bulk properties \cite{Onsager1936-wf}. Among these models, the \acl{PCM} (\acs{PCM})
is the most widespread and successful \cite{Tomasi2005-vm, Mennucci2008-xl}.

The \acs{PCM} is also an ideal candidate for the creation of a solvation \acs{API}.
Its input from and output to \emph{any} quantum chemistry code is
limited and well-defined.
This provides a natural \acs{API} design: the \acs{API} functions can be compared
\emph{vis-à-vis} with the working equations derived for the different
quantum chemical methods.

We will elucidate the design principles adopted in the development of our
\pcmsolver library: an \emph{open-source} \acs{API} for the inclusion of the
PCM in \emph{any} quantum chemistry code \cite{PCMSolver}.  Use of our
\acs{API} \emph{significantly} limits coding effort on the side of the host:
continuum solvation at the \acs{SCF}
level of theory can be implemented in the host program almost out-of-the-box.  The module is
agnostic of the host code internals and is tested separately from it.

We will emphasize the role of the open-source licensing model and the
importance of the modern workflows adopted in our project.  The adoption
of \git as \acs{DVCS}, \cmake for cross-platform builds and automatic
documentation deployment on \readthedocs simplify extensibility of the
module and promote third-party contributions to the code base, while
continuous integration and nightly testing offer an invaluable level of
confidence in the code.

\begin{figure}[!htb]
  \centering
  \includegraphics[width=.5\textwidth]{pcmsolver-scheme}
\end{figure}

\pcmsolver currently offers the basic PCM functionality already
available in many widespread software packages, such as traditional
collocation solvers for isotropic media.~\autocite{Tomasi2005-vm}
Some unique functionalities are the wavelet Galerkin solvers on smooth
molecular surfaces~\autocite{Weijo2010-hy, Bugeanu2015-tp}, the Green’s
functions for spherical diffuse interfaces~\autocite{DiRemigio2016-nn} and
the delayed solvers for real-time \acl{TDSCF}
simulations.~\autocite{Corni2015-pe}

Careful usage of the C++ object-oriented paradigm is the key to this
wide spectrum of functionalities. The library is in itself made of
modules, communicating by means of composition at the outermost
level of the design.
This allows us to mix languages internally: we currently have Fortran
and C modules alongside the main C++ structure.
Library internal classes balance the dynamic polymorphism offered by
inheritance and the static polymorphism offered by template programming.
~\autocite{Alexandrescu2001-bp, Sutter2004-nt, Langr2012-js}
Preeminent use of composition over inheritance keeps the coupling
between different modules as low as possible.~\autocite{Gamma1995-fd}

References for the various codes we have created interfaces to.
\psicode~\autocite{Turney2012-de}
\DIRAC~\autocite{DIRAC15}
\DALTON~\autocite{DALTON16, Aidas2013-rp}
\LSDALTON~\autocite{LSDALTON16, Aidas2013-rp}
\ReSpect~\autocite{ReSpect citation}
KOALA~\autocite{Hofener2014-ex, Hofener2016-qz}

### 2016-08-26 Chapters/ResponseTheory.tex
Clean up and add one chapter (to be written)
Work on the summary of papers

%*****************************************
\chapter{Response theory and molecular properties}\label{ch:molprop}
%*****************************************

\epigraph{
White light, white light goin' messin' up my brain

White light, aww white light it's gonna drive me insane
}{
--- \textsc{The Velvet Underground}
}

\epigraph{What's the frequency, Kenneth? [...]

@@

%*****************************************
%*****************************************
%*****************************************
%*****************************************
%*****************************************

>>>>

\chapter{Response theory and molecular properties}\label{ch:molprop}

\epigraph{
White light goin' messin' up my brain

White light it's gonna drive me insane
}{
--- \textsc{The Velvet Underground}
}
\epigraph{What's the frequency, Kenneth? [...]

@@

\begin{itemize}
    \item Something about exact state response theory: quasienergy,
      response functions, residues of response functions.~\autocite{}
    \item  Derivation of the linear response function with
      quantum/classical polarizable Hamiltonian and then refer to the
      OpenRSP paper for the open-ended treatment.~\autocite{pcm-openrsp}
\end{itemize}

### 2016-08-26 Chapters/SolvationAndCorrelation.tex
Clean up and add one chapter (to be written)
Work on the summary of papers

% mainfile: ../RobertoDiRemigioPhDThesis.tex
%************************************************
\chapter{A New Perspective on Electron Correlation and Solvation}\label{ch:solvation-correlation}

@@

The analyses and derivations in this Chapter are preeminently based
on the \acl{CC} wave function Ansatz.
We will assume a closed-shell, \acs{HF} single reference function and
construct our correlated wave function as an exponential mapping:
\begin{equation}
  \ket{\mathrm{CC}} = \expo{T}\ket{\mathrm{HF}}.
\end{equation}
The \emph{cluster} operator appearing in the exponential is given as:
\begin{equation}
  T = \sum_{p=1}^\mathcal{M} T_p = \sum_{p=1}^\mathcal{M}\sum_{\mu_p}\tamp{\mu_p}\cluster{\mu_p}
\end{equation}
that is, as a truncated sum of excitation operators, $\cluster{\mu_p}$,
times the corresponding cluster amplitudes, $\tamp{\mu_p}$.
Here $\mu_p$ is the $\mu$-th excitation at the $p$-th excitation level and
$\mathcal{M}$ is the truncation level.
One of the main strengths of the \acs{CC} model is its size extensivity
which stems directly from the exponential nature of the wave
operator.~\autocite{Shavitt, Helgaker2000-tz}

>>>>

% mainfile: ../RobertoDiRemigioPhDThesis.tex
%************************************************
\chapter{Electron Correlation and Solvation}\label{ch:solvation-correlation}

@@

The analyses and derivations in this Chapter are preeminently based
on the \acl{CC} wave function Ansatz.
We will assume a closed-shell, \acs{HF} single reference function and
construct our correlated wave function as an exponential mapping:
\begin{equation}
  \ket{\mathrm{CC}} = \expo{T}\ket{\mathrm{HF}}.
\end{equation}
The \emph{cluster} operator appearing in the exponential is given as:
\begin{equation}
  T = \sum_{p=1}^\mathcal{M} T_p = \sum_{p=1}^\mathcal{M}\sum_{\mu_p}\tamp{\mu_p}\cluster{\mu_p}
\end{equation}
that is, as a truncated sum of excitation operators, $\cluster{\mu_p}$,
times the corresponding cluster amplitudes, $\tamp{\mu_p}$.
Here $\mu_p$ is the $\mu$-th excitation at the $p$-th excitation level and
$\mathcal{M}$ is the truncation level.
One of the main strengths of the \acs{CC} model is its size extensivity
which stems directly from the exponential nature of the wave
operator.~\autocite{Shavitt2009-mr, Helgaker2000-tz}
